\Ace: Growing A Statically-Typed Language Inside a Python

### 2014-03-23 ace-oopsla14/ace-oopsla14.tex
ace-oopsla14 commit

%\documentclass[12pt]{article}
\documentclass[9pt,preprint]{sigplanconf}
% The following \documentclass options may be useful:
%
% 10pt          To set in 10-point type instead of 9-point.
% 11pt          To set in 11-point type instead of 9-point.
% authoryear    To obtain author/year citation style instead of numeric.
\usepackage{listings}
\usepackage[usenames,dvipsnames]{color}
\usepackage{mathpartir}
\usepackage{amsmath}
\usepackage{amssymb} 
\usepackage{amsthm}
\usepackage{ stmaryrd }
\usepackage{verbatimbox}

\usepackage{alltt}
\renewcommand{\ttdefault}{txtt}
\usepackage{color}
\usepackage[usenames,dvipsnames,svgnames,table]{xcolor}
\definecolor{mauve}{rgb}{0.58,0,0.82}
\definecolor{light-gray}{gray}{0.75}
\usepackage{listings}

@@

\lstset{
  language=Python,
  showstringspaces=false,
  formfeed=\newpage,
  tabsize=4,
  commentstyle=\itshape,
  basicstyle=\ttfamily\scriptsize,
  morekeywords={lambda, self, assert, as},
  numbers=left,
  numberstyle=\scriptsize\color{light-gray}\textsf,
  xleftmargin=2em,
  stringstyle=\color{mauve}
}
\lstdefinestyle{Bash}{
    language={}, 
    numbers=left,
    numberstyle=\scriptsize\color{light-gray}\textsf,
    moredelim=**[is][\color{blue}\bf\ttfamily]{`}{`},
}
\lstdefinestyle{OpenCL}{
	language=C++,
	morekeywords={kernel, __kernel, global, __global, size_t, get_global_id, sin, printf, int2}
}

@@

\title{\Ace: Growing a Statically-Typed Language Inside a Python}

\authorinfo{~}{~}{~}
%\authorinfo{Cyrus Omar\and Jonathan Aldrich}
 %          {School of Computer Science\\
  %          Carnegie Mellon University}
   %        {\{comar, jonathan.aldrich\}@cs.cmu.edu}

\maketitle
\begin{abstract}
%Evidence suggests that programmers are reluctant to adopt new languages to
%gain access to new abstractions, even when they agree that these abstractions
%could be valuable. This suggests a need for languages that satisfy two key
%growth criteria: compatibility with existing programming ecosystems; and
%internal extensibility, so that new abstractions do not require new
%languages.
%
%We introduce Ace, a language compatible with the popular Python ecosystem.
%While Python, like other similar languages, is satisfactory for simple
%scripting, Ace is designed for more complex situations where static
%typechecking is beneficial. Unlike most statically-typed languages, Ace's type
%system can be extended from within, by associating compile-time functions with
%type definitions. The compiler selectively invokes these according to a
%type-directed protocol, avoiding interference problems. Despite these design
%constraints, active types in Ace can express the static and dynamic semantics
%of a range of functional, object-oriented, parallel and domain-specific
%abstractions, all within libraries.

Evidence suggests that programmers are reluctant to adopt new programming languages to gain access to new abstractions, even when they agree that these abstractions would be valuable to them. This suggests a need for languages that are  \emph{compatible} with existing languages, tools and infrastructure and \emph{internally extensible}, so that adopting a new  primitive abstraction requires only importing a library in the usual way.

In this paper, we introduce Ace, a language compatible with tools and infrastructure developed for Python, one of the most  widely-adopted dynamically-typed languages today. While Python, like other similar languages, was designed for simple scripting tasks, Ace is designed for more complex situations where static typechecking and programmatic control over compilation may be beneficial. Unlike most statically-typed languages, however, Ace's type system and semantics can be extended from within by novel mechanisms that avoid key interference issues faced by previous mechanisms.  
We show that these can be used to implement a range of statically-typed functional, object-oriented, parallel, low-level and domain-specific abstractions, as well as safe interoperability layers with existing languages, as orthogonal libraries. 
%Ace can be used from the shell or interactively from within Python\todo{take this sentence out?}.
%Researchers developing languages and abstractions for high-performance computing must consider a number of design criteria, including performance, verifiability, portability and ease-of-use. Despite the deficiencies of legacy tools and the availability of seemingly superior options, end-users have been reluctant to adopt new language-based abstractions. We argue that this can be largely attributed to a failure to consider three additional criteria: continuity, extensibility\- and interoperability. This paper introduces Ace, a language that aims to satisfy this more comprehensive set of design criteria. To do so, Ace introduces several novel compile-time mechanisms, makes principled design choices, and builds upon existing standards in HPC, particularly Python and OpenCL. OpenCL support, rather than being built into the language, is implemented atop an extensibility mechanism that also admits abstractions drawn from other seemingly disparate paradigms. The core innovation underlying this and other features of Ace is a novel reification of types as first-class objects at compile-time, representing a refinement to the concept of active libraries that we call \emph{active types}. We validate our overall design by considering a case study of a simulation framework enabling the modular specification and efficient execution of ensembles of neural simulations across a cluster of GPUs.
\end{abstract}

%\category{D.3.2}{Programming Languages}{Language Classifications}[Extensible Languages]
%\category{D.3.4}{Programming Languages}{Processors}[Compilers]
%\category{F.3.1}{Logics \& Meanings of Programs}{Specifying and Verifying and Reasoning about Programs}[Specification Techniques]
%\keywords
%type-level computation, typed compilation
\section{Introduction}\label{intro}
Asking programmers to import a new library is far more practical than asking them to  adopt a new programming language. Indeed, recent empirical studies underscore  the difficulties of driving new languages into adoption, finding that extrinsic factors like compatibility with existing codebases and libraries, team familiarity and tool support are at least as important as intrinsic factors \cite{Meyerovich:2013:EAP:2509136.2509515,chen05,nguyen2010survey}. As a result, many developers cannot use  abstractions they might prefer because these abstractions are only available in languages they cannot adopt \cite{Meyerovich:2012:SDR:2414721.2414724,Meyerovich:2013:EAP:2509136.2509515}. This issue was perhaps most succinctly expressed by a participant in a recent study by Basili et al. \cite{basili2008understanding} who stated ``I hate MPI, I hate C++. [But] if I had to choose again, I would probably choose the same.''

Unfortunately, researchers and domain experts who design and develop potentially useful new abstractions can find it difficult to implement them in terms of the general-purpose abstraction mechanisms available in mainstream languages. This is particularly salient for abstractions that require support from the typechecker or compiler, such as those focused on correctness and performance, as well as those that  introduce more concise or natural notations for existing abstractions. For example, a recent controlled study comparing a new language, Habanero-Java (HJ), with a comparable library, \verb|java.util.concurrent|, found that the language-based approach was more concise, correct and easy-to-use, but concluded that the library-based approach was more practical outside the classroom because HJ introduced new constructs and keywords into Java, requiring the adoption of a new toolchain, which could lead to compatibility issues with plain Java code \cite{cave2010comparing}.

Internally-extensible languages promise to reduce the need for new standalone languages by giving abstraction providers more direct control over a base language's syntax and semantics from within libraries. By using such a language, programmers  gain the ability to granularly import the primitive abstractions best suited to each part of their application or library. The research community thus gains the ability to more easily develop, deploy and evaluate new abstractions in the context of existing codebases, narrowing one of the gaps between research and practice \cite{basili2008understanding}.

%Indeed, abstractions could compete on their individual merit, rather than  by the serious adoption barriers mentioned above.

Unfortunately, internally-extensible languages available today have several problems. First, an extension mechanism itself may require modifying a base language with constructs for defining, importing and using extensions. The extension mechanism is not itself a library, so it faces a ``bootstrapping'' problem: it must overcome many of the same extrinsic issues as other new languages like HJ. The extension mechanisms available today also have some intrinsic problems related to safety and expressiveness that require technical solutions before it would be appropriate to widely rely on them.

This paper describes the design and implementation of Ace, an internally-extensible language designed considering both extrinsic and intrinsic criteria. To solve the bootstrapping problem, Ace is implemented entirely as a library within the popular Python programming language. Ace and Python share a common syntax and package system, allowing Ace to leverage its well-established tools and infrastructure directly. Python serves as the compile-time metalanguage for Ace, but Ace functions themselves do not operate according to Python's fixed dynamically-typed semantics  (cf. \cite{Politz:2013:PFM:2509136.2509536,python}). Instead, Ace has a statically-typed semantics that can be extended by users from within libraries.

More specifically, each Ace function can be annotated with a base semantics that determines the meaning of simple expressions like literals and certain statements. The semantics of the remaining expressions and statements are governed by logic associated with the type of a designated subexpression. We call the user-defined base semantics \emph{active bases} and the types in Ace \emph{active types}, borrowing terminology from \emph{active libraries} (\cite{activelibraries}, see Sec. \ref{related}). Both are objects that can be defined and manipulated at compile-time using Python. An important consequence of this mechanism is that it permits \emph{compositional} reasoning -- active bases and active types govern only specific non-overlapping portions of a program. As a result, clients are able to import any combination of extensions with the confidence that link-time ambiguities cannot occur (unlike many previous approaches, as we discuss in Sec. \ref{related}).

The \emph{target} of compilation is also user-defined. We will show examples of Ace targeting Python as well as OpenCL, CUDA and C99, lower-level languages often used to program hardware. An active base or type can support multiple \emph{active targets}, which mediate translation of Ace code to code in a target language. Ace functions targeting a language with Python bindings can be called directly from Python scripts, with compilation occurring implicitly. For some data structures, types can propagate from Python into Ace. We show how this can be used to streamline the kinds of interactive workflows that Python is often used for. Ace can also be used non-interactively from the shell, producing source files that can be further compiled and executed by external means.

The remainder of the paper is organized as follows: in Sec. \ref{usage}, we describe the basic structure and usage of Ace with an example library that internalizes and extends the OpenCL language. Then in Sec. \ref{att}, we show how this and other libraries are implemented by detailing the extension mechanisms within Ace. To explain and demonstrate the expressiveness of these mechanisms (in particular, active types) further, we continue in Sec.  \ref{examples} by showing a diverse collection of abstractions drawn from different language paradigms that can be implemented as orthogonal libraries in Ace. We include functional datatypes, objects, and several other abstractions. In Sec.  \ref{related}, we compare Ace to related work on language extensibility and metaprogramming. We conclude in Sec. \ref{discussion} by summarizing our contributions, discussing the essential features needed by a host language to support these mechanisms, and describing their limitations and potential future work.

\section{Language Design and Usage}\label{usage}
Listing \ref{map} shows an example of an Ace file. As promised, the top level of an Ace file is written directly in Python, requiring no modifications to the language (versions 2.6+ or 3.3+) nor features specific to CPython (so Ace supports alternative implementations like Jython, IronPython and PyPy). This choice pays immediate dividends on line 1: Ace's package system is Python's package system, so Python's build tools (e.g. \verb|pip|) and package repostories (e.g. \verb|PyPI|) are directly available for distributing Ace libraries.

The top-level statements in an Ace file, like the \verb|print| statement on line 10, are executed to control the compile-time behavior, rather than the run-time behavior, of the program. That is, Python serves as the \emph{compile-time metalanguage} (and, as we will see shortly, the \emph{type-level language}) of Ace. %For readers familiar with C/C++, Python can be thought of as serving a role similar to (but more general than) its preprocessor and template system (as we will see).
Functions containing run-time behavior, like \verb|map|, are governed by a semantics that can differ from Python's (in ways that we will describe below), but they share Python's syntax. As a consequence, users of Ace  immediately benefit also from an ecosystem of well-developed tools that work with Python syntax, including parsers, code highlighters, editor modes, style checkers and documentation generators.

\subsection{OpenCL as an Active Library}
The code in this section uses \verb|clx|, an example library implementing the semantics of the OpenCL programming language and extending it with some additional useful types, which we will discuss shortly. Ace itself has no built-in support for OpenCL.

@@

\subsection{Generic Functions}\label{genfn}
\begin{codelisting}
\lstinputlisting[linebackgroundcolor={\btLstHL{5-8}}]{listing3.py}
\caption{[\texttt{listing\ref{map}.py}] A generic data-parallel higher-order map function targeting OpenCL.}
\label{map}
\end{codelisting}
\begin{codelisting}[t]
\lstinputlisting[commentstyle=\color{mauve}]{listing7.py}
\caption{[\texttt{listing\ref{metaprogramming}.py}] Metaprogramming with Ace, showing how to construct generic functions from abstract syntax trees.}
\label{metaprogramming}
\end{codelisting}
Lines 3-4 introduce \verb|map|, an Ace function of three arguments that is governed by the \emph{active base} referred to by \verb|clx.base| and targeting the \emph{active target} referred to by \verb|clx.opencl|. The active target determines which language the function will compile to (here, the OpenCL kernel language) and mediates code generation.

>>>>

%\documentclass[12pt]{article}
\documentclass[10pt,preprint]{sigplanconf}
\newcommand{\lamAce}{\lambda_{\text{Ace}}}
% The following \documentclass options may be useful:
%
% 10pt          To set in 10-point type instead of 9-point.
% 11pt          To set in 11-point type instead of 9-point.
% authoryear    To obtain author/year citation style instead of numeric.
\usepackage{listings}
\usepackage[usenames,dvipsnames]{color}
\usepackage{mathpartir}
\usepackage{amsmath}
\usepackage{amssymb} 
\usepackage{amsthm}
\usepackage{ stmaryrd }
\usepackage{verbatimbox}

\usepackage{alltt}
\renewcommand{\ttdefault}{txtt}
\usepackage{color}
\usepackage[usenames,dvipsnames,svgnames,table]{xcolor}
\definecolor{mauve}{rgb}{0.58,0,0.82}
\definecolor{light-gray}{gray}{0.5}
\usepackage{listings}

@@

\lstset{
  language=Python,
  showstringspaces=false,
  formfeed=\newpage,
  tabsize=4,
  commentstyle=\itshape\color{light-gray},
  basicstyle=\ttfamily\scriptsize,
  morekeywords={lambda, self, assert, as,type},
  numbers=left,
  numberstyle=\scriptsize\color{light-gray}\textsf,
  xleftmargin=2em,
  stringstyle=\color{mauve}
}
\lstdefinestyle{Bash}{
    language={}, 
    numbers=left,
    numberstyle=\scriptsize\color{light-gray}\textsf,
    moredelim=**[is][\color{blue}\bf\ttfamily]{`}{`},
}
\lstdefinestyle{OpenCL}{
	language=C++,
	morekeywords={kernel, __kernel, global, __global, size_t, get_global_id, sin, printf, int2}
}

@@

\title{\Ace: Growing A Statically-Typed Language Inside a Python}

\authorinfo{~}{~}{~}
%\authorinfo{Cyrus Omar\and Nathan Fulton\and Jonathan Aldrich}
 %          {School of Computer Science\\
  %          Carnegie Mellon University}
   %        {\{comar, nfulton, aldrich\}@cs.cmu.edu}

\maketitle
\begin{abstract}
%Evidence suggests that programmers are reluctant to adopt new languages to
%gain access to new abstractions, even when they agree that these abstractions
%could be valuable. This suggests a need for languages that satisfy two key
%growth criteria: compatibility with existing programming ecosystems; and
%internal extensibility, so that new abstractions do not require new
%languages.
%
%We introduce Ace, a language compatible with the popular Python ecosystem.
%While Python, like other similar languages, is satisfactory for simple
%scripting, Ace is designed for more complex situations where static
%typechecking is beneficial. Unlike most statically-typed languages, Ace's type
%system can be extended from within, by associating compile-time functions with
%type definitions. The compiler selectively invokes these according to a
%type-directed protocol, avoiding interference problems. Despite these design
%constraints, active types in Ace can express the static and dynamic semantics
%of a range of functional, object-oriented, parallel and domain-specific
%abstractions, all within libraries.
Programmers are justifiably reluctant to adopt new language dialects to access stronger type systems. This suggests a need for a language that is \emph{compatible} with existing libraries, tools and infrastructure and that has an \emph{internally extensible type system}, so that adopting and combining type systems requires only importing libraries in the usual way, without the need to resolve link-time ambiguities.

We introduce Ace, an extensible statically typed language embedded within and compatible with Python, a widely-adopted dynamically typed language. Python serves as Ace's type-level language. Rather than building in a particular type system, the Ace compiler inverts control over type assignment and translation to type-level functions according to a type-directed protocol that does not lead to ambiguities when type systems are combined. We show that this protocol is flexible enough to admit library-based implementations of types  drawn from functional, object-oriented, parallel, low-level and domain-specific languages. We also show how this permits the construction of safe, natural and extensible foreign function interfaces using a novel form of ``staged'' type propagation from Python into Ace. We also give a simplified calculus that describes how a type-level language can be used to enable type system extensions and demonstrates how type safety can be maintained by lifting a technique from the typed compilation literature into the language.%To show that the method scales, we give a full implementation of the OpenCL programming language as a library. %We briefly examine the type theoretic foundations of this approach and give modular criteria that guarantee that an extension is safe and modular. %Ace can be used from the shell or interactively from within Python\todo{take this sentence out?}.
%Researchers developing languages and abstractions for high-performance computing must consider a number of design criteria, including performance, verifiability, portability and ease-of-use. Despite the deficiencies of legacy tools and the availability of seemingly superior options, end-users have been reluctant to adopt new language-based abstractions. We argue that this can be largely attributed to a failure to consider three additional criteria: continuity, extensibility\- and interoperability. This paper introduces Ace, a language that aims to satisfy this more comprehensive set of design criteria. To do so, Ace introduces several novel compile-time mechanisms, makes principled design choices, and builds upon existing standards in HPC, particularly Python and OpenCL. OpenCL support, rather than being built into the language, is implemented atop an extensibility mechanism that also admits abstractions drawn from other seemingly disparate paradigms. The core innovation underlying this and other features of Ace is a novel reification of types as first-class objects at compile-time, representing a refinement to the concept of active libraries that we call \emph{active types}. We validate our overall design by considering a case study of a simulation framework enabling the modular specification and efficient execution of ensembles of neural simulations across a cluster of GPUs.
\end{abstract}

%\category{D.3.2}{Programming Languages}{Language Classifications}[Extensible Languages]
%\category{D.3.4}{Programming Languages}{Processors}[Compilers]
%\category{F.3.1}{Logics \& Meanings of Programs}{Specifying and Verifying and Reasoning about Programs}[Specification Techniques]
%\keywords
%type-level computation, typed compilation
\section{Introduction}\label{intro}
Asking programmers to import a new library is simpler than asking them to adopt a new programming language. Indeed, recent empirical studies underscore the difficulties of driving languages into adoption, finding that extrinsic factors like compatibility with large existing code bases,  library support, team familiarity and tool support are at least as important as intrinsic features of the language  \cite{Meyerovich:2013:EAP:2509136.2509515,chen05,nguyen2010survey}.

Unfortunately, researchers and domain experts (\emph{providers}) designing potentially useful new abstractions can sometimes find it difficult to implement them as libraries, particularly when they require strengthening a language's type system. In these situations, abstraction providers often develop a new language or language dialect. Unfortunately, this \emph{language-oriented approach} \cite{language-oriented} does not scale to large applications: using  components written in languages with different type systems can be awkward and lead to safety problems and performance overhead at interface boundaries. %To avoid this, it is necessary to combine the different languages and their associated tools, but this often requires significant effort and expertise (we will discuss this further in Sec. \ref{related-work}).

This is a problem even when a dialect introduces only a small number of new constructs. For example, a recent  study \cite{cave2010comparing} comparing a Java dialect, Habanero-Java (HJ), with a comparable library, \verb|java.util.concurrent|, found that the language-based abstractions in HJ were easier to use and provided useful static guarantees. Nevertheless, it concluded that the library-based abstractions remained more practical outside the classroom because HJ, as a distinct dialect of Java with its own type system, would be difficult to use in settings where some developers had not adopted it, but needed to interface with code that had adopted it. It would also be difficult to use it in combination with other abstractions also implemented as dialects of Java. Moreover, its tool support is more limited. 
This suggests that today, programmers and development teams cannot use  the abstractions they might prefer because they are only available bundled with languages they cannot adopt \cite{Meyerovich:2012:SDR:2414721.2414724,Meyerovich:2013:EAP:2509136.2509515}.

 %This issue was perhaps most succinctly expressed by a participant in a recent study by Basili et al. \cite{basili2008understanding} who stated ``I hate MPI, I hate C++. [But] if I had to choose again, I would probably choose the same.''
%Unfortunately, researchers and domain experts (collectively, \emph{providers}) who design potentially useful new abstractions often find it difficult to implement them in terms of the general-purpose abstraction mechanisms available in existing languages. 
%Using a foreign construct can be  unnatural and, if the languages have  different type systems, can also cause safety problems.

Internally extensible languages promise to reduce the need for new dialects by giving abstraction providers more direct control over a language's syntax and static semantics from within libraries. %In   an internally extensible language, programmers  can \emph{granularly} import the primitive constructs best suited to each component of their application or library. Providers can thus more easily develop, deploy and evaluate new abstractions in the context of existing codebases, narrowing one of the gaps between research and practice \cite{basili2008understanding}. 
%Indeed, abstractions could compete on their individual merit, rather than  by the serious adoption barriers mentioned above.
Unfortunately, the mechanisms available today have several problems. First, they are themselves often available only within a dialect of an existing language and thus face a ``chicken-and-egg'' problem: a language like SugarJ \cite{sugarj} must overcome the same extrinsic issues as a language like HJ. Second, giving providers too much control over the language can lead to intrinsic safety issues, ambiguities and conflicts between extensions, as we will discuss. Too little control, on the other hand, leaves it difficult to implement interesting abstractions.%It is difficult to guarantee that extensions are sound in isolation and more so to guarantee that they can be {safely} combined. %This requires further research before it would be appropriate to widely rely on them (which we will discuss as we go on).

This paper describes the design and implementation of Ace, an internally extensible, statically typed programming language designed considering both these extrinsic and intrinsic factors. To address the ``chicken-and-egg'' problem, Ace is implemented within Python, entirely as a library  \cite{Politz:2013:PFM:2509136.2509536,python}. The top-level of an Ace file can be thought of as a \emph{compilation script} written in Python and, as we will discuss, Python serves as Ace's type-level language.  Ace and Python share a common syntax and package system, so users of Ace can leverage established tools and infrastructure directly. %The semantics of  define run-time behavior.

Functions marked by a decorator as under the control of Ace are \emph{statically typechecked}. This makes it possible to rule out many potential issues ahead of execution and thus avoid the need for costly dynamic checks. % Instead, Ace has a statically-typed semantics that can be extended by users from within libraries. 
However, there is \emph{no particular type system built into Ace}. Instead, a type is any  {compile-time object} implementing the \verb|ace.Type| interface. We call these \emph{active types}. When statically assigning a type to a compound form, e.g. \verb|e.x|, the Ace {compiler} delegates to the type of a subexpression, e.g. to the type of \verb|e|, to determine how to assign a type to the expression as a whole. As we will show, this {type-directed protocol}, which we call \emph{active type assignment}, permits the expression of a rich variety of type systems as libraries. Unlike syntax-directed protocols, there cannot be ambiguities when separately defined type sytems are combined (extensions have exclusive control over non-overlapping sets of expressions by construction). Base cases, e.g.  variables and literals, are handled on a {per-function basis} by a \emph{base semantics}, also a compile-time object implementing an interface, \verb|ace.Base|.

% As a result, clients are able to import any combination of extensions with the confidence that link-time ambiguities cannot occur .

The dynamic behavior of an Ace function is determined by translation to a \emph{target language}. We begin by  targeting Python, then discuss targeting OpenCL and CUDA, lower-level languages used to program many-core processors (e.g. GPUs). Active types and bases control translation by the same type-directed protocol that governs type assignment. We call this phase \emph{active translation}. A \emph{target} mediates this process and is also a user-defined compile-time object implementing an interface, \verb|ace.Target|. 
%One base or type can support multiple targets. 
When targeting a typed language, care must be taken to ensure that well-typed Ace expressions have well-typed translations, so we integrate a technique for compositionally reasoning about compiler correctness (developed for the TIL compiler for Standard ML \cite{TIL}) into the language.

Ace can be used from the shell, producing source files that can be further compiled or executed by external means. Ace functions targeting a language with Python bindings (e.g. Python itself, as well as the others just described) can also be compiled and invoked interactively from Python scripts. Ace supports a form of \emph{ad hoc polymorphism} (based on singleton types) whereby types propagate into generic functions. This occurs both between Ace functions and when calling Ace functions from Python. In the latter case, the ``dynamic type'' of a Python value determines a static type ``just-in-time'' when a call occurs. This represents a form of implicit \emph{staging} and significantly streamlines interactive workflows (common in, for example, scientific computing) that alternate between Python for orchestration and productivity-oriented tasks (e.g. exploratory plotting) and an external language via a foreign-function interface (FFI) for performance- and correctness-critical portions.%We will discuss in detail as our primary realistic case study.

The remainder of the paper is organized as follows: in Sec. \ref{usage}, we describe the basic structure and usage of Ace with an example that simultaneously uses dynamically typed terms (which can be thought of as having the static type \verb|dyn|) alongside statically typed immutable records and a simple prototypic object system, all implemented orthogonally as active types. In Sec. \ref{att}, we detail how active type assignment and translation delegate control, at compile-time, to active types and bases. In Sec. \ref{theory}, we reduce this mechanism to  a core lambda calculus, $\lamAce$,  clarify the relationship to   type-level computation and show how type safety is maintained. In Sec. \ref{targets}, we discuss generic functions, FFIs and staging via a complete implementation of the OpenCL type system (and extensions to it) as a library.  In Sec.  \ref{related}, we compare Ace to related work. We conclude in Sec. \ref{discussion} by summarizing our contributions, highlighting the key features needed by a host language to support these mechanisms, and discussing limitations and potential future work. The appendix contains details omitted from the main body and additional examples.

\section{Language Design and Usage}\label{usage}

%
Listing \ref{example} shows an example of an Ace compilation script. As promised, compilation scripts are written directly in Python. Ace requires no modifications to the language or features specific to the primary implementation, CPython (so Ace supports alternative implementations like Jython and PyPy). This choice pays immediate dividends on the first five lines: Ace's package system is Python's package system, so Python's build tools (e.g. \verb|pip|) and package repostories (e.g. \verb|PyPI|) are directly available for distributing libraries that use Ace. Note that all of the type systems that we discuss are implemented entirely as  libraries.

%The top-level statements in an Ace file, like the \verb|print| statement on line 10, are executed to control the compile-time behavior, rather than the run-time behavior, of the program. %That is, Python serves as the \emph{compile-time metalanguage} (and, as we will see shortly, the \emph{type-level language}) of Ace. %For readers familiar with C/C++, Python can be thought of as serving a role similar to (but more general than) its preprocessor and template system (as we will see).

\subsection{Types}
\begin{codelisting}
\begin{lstlisting}
from examples.py import py, string
from examples.fp import record
from examples.oo import proto
from examples.num import decimal
from examples.regex import string_in

A = record['amount' : decimal[2]]
C = record[
  'name' : string, 
  'account_num' : string_in[r'\d{10}'],
  'routing_num' : string_in[r'\d{2}-\d{4}/\d{4}']]
Transfer = proto[A, C]

@py
def log_transfer(t):
  """Logs a transfer to the console."""
  {t : Transfer}
  print "Transferring %s to %s." % (
    [string](t.amount), t.name)

@py
def __toplevel__():
  print "Hello, run-time world!"
  common = {name: "Annie Ace", 
            account_num: "0000000001", 
            routing_num: "00-0000/0001"} (C)
  t1 = ({amount: 5.50}, common) (Transfer)
  t2 = ({amount: 15.00}, common) (Transfer)
  log_transfer(t1)
  log_transfer(t2)

print "Hello, compile-time world!"
\end{lstlisting}
\caption{[\texttt{listing\ref{example}.py}] An Ace compilation script.}
\label{example}
\end{codelisting}

Types are constructed programmatically during execution of the compilation script. This stands in contrast to many  contemporary statically-typed languages, where types (e.g. datatypes, classes, structs) can only be declared. Put another way, Ace supports \emph{type-level computation} \cite{tlc} and Python is its type-level language (discussed further in Sec. \ref{theory}). 
%The compilation script can assign shorter names to types for convenience (removing the need for  facilities like \verb|typedef| in C or \verb|type| in Haskell). 
In our example, we see several types being constructed:
\begin{enumerate}
\item On line 7, we construct a functional record type with a single (immutable) field named \verb|amount| with type \verb|decimal[2]|, which classifies decimal numbers with two decimal places. Here, \verb|record| and \verb|decimal| are \emph{indexed type constructors}. We will see how they are implemented in Sec. \ref{att}, but note that syntactically, \verb|record| borrows Python's syntax for array slices (e.g. \verb|a[min:max]|). Note that the field name could equivalently have been written \lstinline{'a' + 'mount'}, again emphasizing that types and indices are constructed programatically by a Python script.
\item On lines 8-11, we construct another record type. The field \verb|name| has type \verb|string|, defined in \verb|examples.py|, while the fields \verb|account_num| and \verb|routing_num| have more interesting types, classifying strings guaranteed statically to be in the language of a regular expression  (written, to avoid needing to escape backslashes, using Python's \emph{raw string literals}). Our appendix specifies the typechecking rules, based on \cite{nfulton-thesis}. We include it to emphasize that we aim to support specialized type systems,  not just general purpose constructs like records and numbers.
\item Prototypic inheritance is a feature of some dynamically-typed languages (most notably, Self \cite{self} and Javascript). On line 12, we consider a simple statically-typed variant. The type \verb|Transfer| classifies terms consisting of a \todo{what is the right term for this?}\emph{fore} of type \verb|A| and a \emph{prototype} of type \verb|C|. If a field cannot be found in the fore, the type system will delegate (here, statically) to the prototype. This makes it easy to share the values of the fields of a common prototype amongst many fores. In our example, we will perform multiple transfers differing only in amount.
\end{enumerate}
\subsection{Functions}
Functions implementing run-time behavior are distinguished by a decorator specifying their  \emph{base semantics} (here \verb|py| from \verb|examples.py|; we will discuss this further below). These functions are, however, still written using Python's syntax, a choice that is again valuable for extrinsic reasons: users of Ace can use a variety of  tools designed to work with Python source code  without  modification, including code highlighters (like the one used in generating this paper), editor plugins, style checkers and documentation generators. We will see how, with a bit of cleverness, Python's syntax can support a variety of static semantics.
\begin{codelisting}
\begin{lstlisting}[style=Bash]
$ `acec listing1.py`
Hello, compile-time world!
[acec] _listing1.py successfully generated.
$ `python _listing1.py`
Hello, run-time world!
Transferring 5.50 to Annie Ace.
Transferring 15.00 to Annie Ace.
\end{lstlisting}
\caption{Compiling \texttt{listing\ref{example}.py} using \texttt{acec}. Both steps can be performed at once by writing \lstinline[style=Bash]{`ace listing1.py`} (line 3 will not be printed with this command).}
\label{external-compilation}
\end{codelisting}
\begin{codelisting}[t]
\begin{lstlisting}
import examples.num.runtime as __ace_0

def print_transfer(t):
  print "Transferring %s to %s." % (
    __ace_0.decimal_to_str(t[0],2), t[1][0])

print "Hello, run-time world!"
common = ("Annie Ace", "0000000001", "00-0000/0000")
t1 = ((5,50), common)
t2 = ((15,0), common)
print_check(t1)
print_check(t2)
\end{lstlisting}
\caption{[\texttt{\_listing\ref{example}.py}] The file generated in Listing \ref{external-compilation}.}
\label{example-out}
\end{codelisting}

On lines 14-18, we see the function \verb|log_transfer|. As with a Python function, we write a documentation string like the one shown specifying, informally, the behavior of the function. Before moving into the body, however, we  also write a \emph{type signature} (line 17) stating simply that the type of \verb|t| is \verb|Transfer|, the prototypic object type described above. As a result, we can assume that \verb|t.amount| and \verb|t.name| have types \verb|decimal[2]| and \verb|string|, respectively. We will return to the details of \verb|print| and the form \verb|[string](t.amount)|, which performs an explicit conversion, shortly.\todo{when?}

Line 17 borrows Python's syntax for dictionary literals to approximate  conventional notation for type annotations.\footnote{If the right-hand side of every annotation is not a type, it will be an error. In the rare situation where there are no type annotations and the first statement is a dictionary literal, users can use Python's \texttt{pass} keyword to skip the line.} % (and following conventions for documentation strings, which also describe ``types'' at the end). % The ``keyword'' \verb|type| was chosen because it is the name of a built-in function in Python (and thus would not be used in the way shown for any other purpose).
 {In version 3.0 of Python, syntax for annotating arguments with arbitrary metadata using \verb|:| was introduced,  intended for use by future tools (further suggesting that users of dynamically typed languages see potential in a static typing discipline)  \cite{PEPwhatever}. Ace supports both notations when the compilation script is run using Python 3.\textit{x}, but we use the more universally available notation here to satisfy our extrinsic goals: the Python 2.\text{x} series  remains the most widely used by a large margin as of this writing (we will consider cross-compilation in Sec. \ref{targets}, however).

\subsection{Introductory Forms}
On lines 21-30, we define the function \verb|__toplevel__| (the base will consider this a special name for the purposes of compilation, discussed in the Sec. \ref{compilation}). After printing a run-time greeting, we construct a value of type \verb|C| on lines 24-26. Recall that \verb|C| is a record type, so we provide the names of the fields (here, without quotes because we are no longer in the type-level langauge) and values of the appropriate type using syntax for a dictionary literal. Because there may be multiple types that use this form, we say which type the literal should be \emph{analyzed} against with the \emph{literal ascription}  \verb|(C)|. This is another repurposing of Python's syntax: here function application. When the ``function'' has one of Python's literal forms and the (single) argument is a type, we treat it instead as an \emph{ascribed introductory form}. The literal is treated as an introductory form for that type (we will see precisely how this occurs shortly). We see the same form used with tuple literals on lines 27-28 to introduce terms of type \verb|Transfer|.

\subsection{Bidirectional Typechecking}
The string, number and dictionary forms inside the outermost forms on lines 24-28 are also introductory forms, but we do not include an ascription because the definition of \verb|C| determines which types they should be analyzed against. As we will discuss shortly, Ace has a \emph{bidirectional type system} distinguishing locations where a type must be \emph{synthesized} from an expression (e.g. at the top-level) and locations where an expression is \emph{analyzed} against a type \cite{LovasPfenning08}. The base determines the type a literal should be analyzed against in a synthetic context. For example, if we did not specify the literal ascription \verb|(C)| on line 26, the base would analyze it against the Python dictionary type, which would cause a type error because the variables used as keys were not defined in the surrounding context.

\subsection{Compilation}\label{compilation}

%Because \verb|t| will be statically guaranteed to have type \verb|Transfer|,  
\subsection{B}
\section{Active Type Synthesis and Translation}\label{att}
The core of Ace consists of about 1500 lines of Python code implementing its primary concepts: generic functions, concrete functions, active types, active bases and active targets.  The latter three comprise Ace's extension mechanism. Extensions provide semantics to, and govern the compilation of, Ace functions, rather than logic in Ace's core. %Indeed, the name ``Ace'' might itself be an acronym: it is an ``active compilation environment''.

\section{Theoretical Foundations}\label{theory}
\section{Generic Functions, Targets and Interactive Execution}\label{targets}
\subsection{OpenCL as an Active Library}
The code in this section uses \verb|clx|, an example library implementing the semantics of the OpenCL programming language and extending it with some additional useful types, which we will discuss shortly. Ace itself has no built-in support for OpenCL.

@@

\subsection{Generic Functions}\label{genfn}
%\begin{codelisting}[t]
%\lstinputlisting[commentstyle=\color{mauve}]{listing7.py}
%\caption{[\texttt{listing\ref{metaprogramming}.py}] Metaprogramming with Ace, showing how to construct generic functions from abstract syntax trees.}
%\label{metaprogramming}
%\end{codelisting}
